Translation of B Implementations to the LLVM: Control and Modularity

### 2013-10-24 translation.tex
Fixed root Python script. Fixed code generation so that LLVM code units generated for B multi-machines projects are linkable.

\begin{figure}
\begin{center}
\includegraphics[width=.8\textwidth]{b-project.pdf}
\caption{Structure of the imports relation in a B project}
\label{fig:structure}
\end{center}
\end{figure}
Figure~\ref{fig:structure} illustrates this with the structure of a B project
where an implementation \B{I0} of a specification \B{M0} imports one unnamed
instance of a basic machine \B{M1} and two instances named \B{A} and \B{B} of a
developed machine \B{M2}. Its implementation \B{I2} in turn imports two
instances \B{C} and \B{D} of a developed machine \B{M3}, implemented as \B{I3}
(which has no imports itself).  The generation of LLVM code for the project
shall produce the representation for one instance of \B{I0}, two instances of
\B{I2} and four instances of \B{I3}.

@@

Before implementing this translation, we want validate the rules that define it.
For this, we follow a pragmatic approach, based on testing and simulation of the
generated LLVM artifacts, and using the existing LLVM tool set.

First, we define a sample set of small B implementation modules, that contain
all the B constructs handled in the translation. For each such module, the rules
are applied manually, yielding a LLVM module. This LLVM module is then amended
with extra code that have two purposes:
\begin{itemize}
\item a standalone executable may be generated from it;
\item the executable creates an instance of the B implementation, applies the
  initialization code, and offers to the user the choice to interactively invoke
  all the operations of the implementation; after initialization and each
  operation application the state of the implementation is printed. 
\end{itemize}

The LLVM tools that are employed are:
\begin{description}
\item[llvm-as] The LLVM assembler generates LLVM bitcode, invoked, e.g., as
  follows:
\begin{verbatim}
llvm-as-mp-3.1 example-01.llvm -o example-01.bc
\end{verbatim}
\item[lli] The LLVM bitcode interpreter, which may be invoked as follows:
\begin{verbatim}
lli-mp-3.1 example-01.bc
\end{verbatim}
\end{description}

Once a rule has been manually validated, we implement it in a prototype tool
written in Python. Unfortunately the tool does not take as input the B0 source
file, but instead a collection of Python definitions corresponding to the
syntactic entities present in the parse tree. For instance, for an integer
concrete variable named \B{value}, we would have the following Python entity:
\begin{verbatim}
value = { "kind": "Vari",
          "id": "value",
          "type": "INT",
          "scope": "impl",
          "root": counter_i}
\end{verbatim}
where \verb'counter_i' would be the Python entity representing the parse tree
root for the B implementation. Again the output generated by the prototype
is compiled and simulated using off-the-shelf LLVM tools \verb'llvm-as' and
\verb'lli'.

>>>>

\begin{figure}
\begin{center}
 \includegraphics[width=.8\textwidth]{B-project.pdf}
\caption{Structure of the imports relation in a B project}
\label{fig:structure}
\end{center}
\end{figure}
Figure~\ref{fig:structure} illustrates this with the structure of a B project
where an implementation \B{I0} of a specification \B{M0} imports one unnamed
instance of a basic machine \B{M1} and two instances named \B{A} and \B{B} of a
developed machine \B{M2}. Its implementation \B{I2} in turn imports two
instances \B{C} and \B{D} of a developed machine \B{M3}, implemented as \B{I3}
(which has no imports itself).  The generation of LLVM code for the project
shall produce the representation for one instance of \B{I0}, two instances of
\B{I2} and four instances of \B{I3}.

@@

Before implementing this translation, we want validate the rules that define it.
For this, we follow a pragmatic approach, based on testing and simulation of the
generated LLVM artifacts, and using the existing LLVM tool set. Consequently,
the rules described in the preceding sections have been prototyped in the Python
programming language. The input of the translation is in a XML-based format,
named BXML, that is generated by the forthcoming version of Atelier-B. For each
B component (machine or implementation), Atelier-B generates a BXML file that
may then be used as input by our prototype. The output of the translation is in
LLVM-IR format. Such output may then be processed by the LLVM tools either to
generate binary in some compatible target platform or LLVM bitcode that may then
be executed using the LLVM interpreter. We used the latter approach to validate
our translation.

The LLVM tools that are employed are:
\begin{description}
\item[llvm-as] The LLVM assembler generates LLVM bitcode, invoked, e.g., as
  follows:
\begin{verbatim}
llvm-as example-01.llvm -o example-01.bc
\end{verbatim}
\item[llvm-link] The LLVM bitcode linker, which may be invoked as follows:
\begin{verbatim}
llvm-link example-01.bc example-02.bc -o prog.bc
\end{verbatim}
\item[lli] The LLVM bitcode interpreter, which may be invoked as follows:
\begin{verbatim}
lli-mp example-01.bc
\end{verbatim}
\end{description}

We describe below the different commands required to use the LLVM framework to
animate a B implementation \B{wd} that uses an instance of a B module
\B{counter}:
\begin{verbatim}
b2llvm counter counter.llvm
b2llvm wd counter.llvm
b2llvm --mode proj wd init-wd.llvm
llvm-as counter.llvm -o counter.bc
llvm-as wd.llvm -o wd.bc
llvm-as init-wd.llvm -o init-wd.bc
llvm-as scaffold-wd.llvm -o scaffold-wd.bc
llvm-link counter.bc wd.bc init-wd.bc scaffold-wd.bc -o anim-wd.bc
lli anim-wd.bc
\end{verbatim}
